Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2021-12-23 analysis/paper/paper.Rmd
Minor edits

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.pos = "!h",
  fig.width = 9,
  fig.path = "../figures/",
  dpi = 300
)
```

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g. @clark2017], which have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.

@@

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project),
          size = 2.25, shape = 21, 
          colour = "black") +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

@@

```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis. Column on the left shows the data as originally compiled, the columns on the right shows how these have been aggregated for the analysis.", fig.height = 7}
# Most of the code for assembling the Sankey plot has been taken directly from 
# various walkthroughs at https://www.r-graph-gallery.com

@@

# Create plot
sn <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              NodeID = "name", Value = "value", colourScale = my_colour,
              sinksRight = FALSE, LinkGroup = "new_categories", 
              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. Some additional configurations and ways to aggregate the variables are also available as supplementary material to the paper.

Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 1 mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].

@@

```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot. Points close together are more similar. By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables. As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns can be made (see Lockyear 2000 for a pedagogical case study).", fig.height = 13}
# Plot sites
plt1 <- fviz_ca_row(calithics, col.row = "black",
                    fill = lithics$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

Thus, while some secondary expectations of the WABI do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself. The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material. Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases [e.g. @bergsvik2001; @bjerck2008; @glørstad2010; @jaksland2001]. To explore this proposition further, these two variables are subjected to more detailed scrutiny below.

@@

The results of the CA appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf. @bjerck2017; @breivik2018; @fuglestvedt2009; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.

@@

The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE. This pattern is evident in both the flint data and when all lithics are treated in aggregate. Furthermore, the increased variation in degree of curation after around 8000 BCE could indicate that these sites were associated with a more varied mobility pattern. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base-camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility [@eigeland2014b]. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. Thus, the degree of expediency for assemblages dominated by non-flint might be somewhat exaggerated when the non-flint material is included, while its exclusion would likely lead to its underestimation. One possible approach could be to weigh the curation index by the proportion of non-flint material in the assemblages. This is not explored further here, however, as the overall tendencies are relatively robust to this effect.

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. The curation index is relatively high and uniform until some time before 8000 BCE. This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility. This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE. Again, this corresponds well with the employed chronological framework. Referring back to the demographic changes that are to take place around this transition, the Microlith phase could thus represent a period where migrating people and new living practices were propagating through societies in south-eastern Norway---a process that in light of the curation data would have concluded around 7000 BCE.

@@

As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. The fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@binford1980]. This indicates, in turn, that the measures represent an empirical link between technological organisation and economic behaviour and mobility patterns [@riel-salvatore2004].

@@

```{r, fig.height = 15, fig.cap = "Same as above, only that the proportion of non-flint is used as a supplementary column. The negligble difference between the plots indicates that the raw material variability is captured by the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated. A) Row plot, B) Column plot."}

@@

splt4 <- fviz_ca_col(nonpropsup, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

@@

```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot."}
all_cats <- sites %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

@@

<!-- \newpage -->

<!-- ### Colophon -->

<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->

<!-- ```{r colophon, cache = FALSE} -->

<!-- # which R packages and versions? -->

<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->

<!-- ``` -->

<!-- The current Git commit details are: -->

<!-- ```{r} -->

<!-- # what commit is this file at?  -->

<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->

<!-- ``` -->

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.pos = "!h",
  fig.width = 9,
  fig.path = "../figures/",
  dpi = 300
)

knitr::opts_knit$set(
  eval.after = "fig.cap"
)
```

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g. @clark2017] and other dimensions that have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.

@@

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project),
          size = 2.25, shape = 21, 
          colour = "black") +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  ggsn::north(sitejoin, scale = 0.2, symbol = 10,
              anchor = c(x = anc[2] - 30000, y = anc[1]) + 5500) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @eigeland2015; @fuglestvedt2007; @mansrud2016]. These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

@@

```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis. The column on the left shows the data as originally compiled. The column on the right shows how these have been aggregated for the analysis.", fig.height = 7}
# Most of the code for assembling the Sankey plot has been taken directly from 
# various walkthroughs at https://www.r-graph-gallery.com

@@

# Create plot
sn <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              NodeID = "name", Value = "value", colourScale = my_colour,
              sinksRight = FALSE, LinkGroup = "new_categories", 
              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 12)

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. A version of the CA using the original artefact categories, as well as some additional configurations and ways to aggregate the variables are also available in the supplementary material to the paper.

Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as the total number of artefacts divided by total excavated m³ as given in the original reports; relative frequency of chips, defined as the proportion of artefacts with size \< 1 mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].

@@

```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot. Points close together are more similar. By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables. As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns across the two plots can be made.", fig.height = 13}
# Plot sites
plt1 <- fviz_ca_row(calithics, col.row = "black",
                    fill = lithics$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

Thus, while some secondary expectations of the WABI as related to cores and blanks do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself. The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material. Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases [e.g. @bergsvik2001; @bjerck2008; @glørstad2010; @jaksland2001]. To explore this proposition further, these two variables are subjected to more detailed scrutiny below.

@@

The results of the CA appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf. @bjerck2017; @breivik2018; @fuglestvedt2009; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatchets and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.

@@

The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE. This pattern is evident in both the flint data and when all lithics are treated in aggregate. Furthermore, the increased variation in degree of curation after around 8000 BCE could indicate that these sites were associated with a more varied mobility pattern. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base-camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility [@eigeland2014b]. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. Thus, the degree of expediency for assemblages dominated by non-flint might be somewhat exaggerated when the non-flint material is included, while its exclusion would likely lead to its underestimation. One possible approach could be to weigh the curation index by the proportion of non-flint material in the assemblages. This is not explored further here, however, as the overall tendencies appear robust to this effect.

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear robust. The curation index is relatively high and uniform until some time before 8000 BCE. This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility. This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE. Again, this corresponds well with the employed chronological framework. Referring back to the demographic changes that are to take place around this transition, the Microlith phase could thus represent a period where migrating people and new living practices were propagating through societies in south-eastern Norway---a process that in light of the curation data would have concluded around 7000 BCE.

@@

As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE. These then remain stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. The fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@binford1980]. This indicates, in turn, that the measures represent an empirical link between technological organisation and economic behaviour and mobility patterns [@riel-salvatore2004].

@@

```{r, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column. The negligble difference between this CA and that above indicates that the raw material variability is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated. These plots thus hide important variability that is captured in the ones presented in the main text. A) Row plot, B) Column plot."}

@@

splt4 <- fviz_ca_col(nonpropsup, col.col.sup = "black", col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

@@

```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
all_cats <- sites %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

@@
